OptiFit: a fast method for fitting amplicon sequences to existing OTUs

### 2021-09-01 paper/paper.Rmd
Incorporate (most of) Pat's revisions
Co-authored-by: Pat Schloss <pschloss@users.noreply.github.com>

<!--
- Go back through https://peerj.com/articles/1487/ to get all the arguments for why the qiime based methods are problematic…
1. The threshold with closed reference is a radius and with de novo its a diameter
2. The gg reference was defined for full length sequences, within a subregion there are duplicate and nearly duplicate sequences
3. Because VSEARCH/USEARCH is a greedy heuristic, its looking for a reference that is close. Although the ordering of the database is not relevant (it was with USEARCH) it gets the wrong number of its (see Figure 5)
-->

@@

Assigning amplicon sequences to operational taxonomic units (OTUs) is often an
important step in characterizing the composition of microbial communities across
large datasets.
OptiClust, a _de novo_ OTU clustering method in the mothur program, has been
shown to produce higher quality OTU assignments than other methods and at
comparable or faster speeds.
A notable difference between _de novo_ clustering and database-dependent methods
is that OTU assignments clustered with _de novo_ methods may change when new
sequences are added to a dataset.
However, in some cases one may wish to incorporate new samples into a previously
clustered dataset without performing clustering again on all sequences, such as
when comparing across datasets or deploying machine learning models where OTUs
are features.
To provide an efficient and robust method to fit amplicon sequence data to
existing OTUs, we developed the OptiFit algorithm.
We tested OptiFit using four microbiome datasets with two different strategies:
by fitting to an external reference database, or by splitting the dataset into a 
reference and query set and fitting the query sequences to the reference set.
The result is a robust implementation of closed and open-reference clustering.
OptiFit produces OTUs of similar quality as OptiClust and at faster speeds when
using the split dataset strategy, although the OTU quality and processing speed
depends on the database chosen when using the external database strategy.
OptiFit provides a suitable option for users who require consistent OTU
assignments at the same quality afforded by _de novo_ clustering methods.

@@

**TODO.**
Advancements in DNA sequencing technology have allowed researchers to affordably
generate millions of reads from microorganisms in diverse natural communities.
allowing us to characterize communities.
\newpage

@@

Amplicon sequencing has become a mainstay of microbial ecology.
Researchers can affordably generate millions of sequences to characterize the
composition of hundreds of samples from microbial communities without the need
for culturing.
In many analysis pipelines, 16S rRNA gene sequences are assigned to
operational taxonomic units (OTUs) to facilitate comparison of taxonomic
composition between communities to avoid the need for classification.
A distance threshold of 3% (or sequence similarity of 97%) is commonly used to
cluster sequences into OTUs based on pairwise comparisons of the sequences
within the dataset.
The method chosen for clustering affects the quality of OTU assignments and thus
may impact downstream analyses of community composition
[@westcott_opticlust_2017; @schloss_application_2016;
@westcott_novo_2015].

There are two main categories of OTU clustering algorithms: _de novo_ and
reference-based.
OptiClust is a _de novo_ clustering algorithm which uses the distance score
between all pairs of sequences in the dataset to cluster them into OTUs by
maximizing the Matthews Correlation Coefficient (MCC)
[@westcott_opticlust_2017].
This approach takes into account the distances between all pairs of sequences
when assigning query sequences to OTUs, in contrast to other _de novo_
methods such as the greedy clustering algorithms implemented in USEARCH and
VSEARCH, which only consider the distance between the query sequence and a
representative centroid sequence in the OTU [@edgar_search_2010;
@rognes_vsearch_2016].
As a result of greedy clustering, some pairs of sequences in the same OTU may
have a greater distance than the specified threshold since only the distance
between each sequence and the centroid sequence is considered while clustering.
In contrast, the OptiClust algorithm enforces that all pairs of sequences must
be within the distance threshold. <!-- i.e. radius vs. diameter -->
A limitation of _de novo_ clustering is that different OTU assignments will be
produced when new sequences are added to a dataset, making it difficult to use
_de novo_ clustering to compare OTUs between different studies.
Additionally, the greedy clustering algorithms are sensitive to the order of the
input sequences: different OTU assignments are produced when the same sequences
are randomly shuffled [@westcott_novo_2015; @he_stability_2015].
Furthermore, since _de novo_ clustering requires calculating and comparing
distances between all sequences in a dataset, the execution time can be slow for
very large datasets.
Reference clustering attempts to overcome the limitations of _de novo_
clustering methods by using a representative set of sequences from a database,
with each reference sequence seeding an OTU.
Commonly, the Greengenes set of representative full length sequences clustered
at 97% similarity is used as the reference with VSEARCH
[@desantis_greengenes_2006; @rognes_vsearch_2016; @noauthor_clustering_nodate].
Query sequences are then assigned to OTUs based on their similarity to the
reference sequences.
Any query sequences that are not within the distance threshold to any of the
reference sequences are either thrown out (closed reference clustering) or
clustered _de novo_ to create additional OTUs (open reference clustering).
While reference-based clustering is generally fast, it is limited by the
diversity of the reference database.
Rare or novel sequences in the sample will be lost in closed reference mode if
they are not represented by a similar sequence in the database.
Previous studies found that the OptiClust _de novo_ clustering algorithm created
the highest quality OTU assignments of all clustering methods [@westcott_opticlust_2017].

To overcome the limitations of current reference-based and _de novo_ clustering
algorithms while maintaining OTU quality, we developed OptiFit, a
reference-based clustering algorithm which uses existing OTUs as the reference
and fits new sequences those reference OTUs.
In contrast to other tools, OptiFit considers all pairwise distance scores
between reference and query sequences when assigning sequences to OTUs in order
to produce OTUs of the highest possible quality.
Here, we tested the OptiFit algorithm with the reference as a database or _de
novo_ OTUs and compared the performance to existing tools.
To evaluate the OptiFit algorithm and compare to existing methods, we used four
published datasets isolated from soil [@johnston_metagenomics_2016], marine
[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
gut [@baxter_microbiota-based_2016] samples. 
OptiFit is available within the mothur software program.

@@

OptiFit begins where OptiClust ends, starting with a list of reference OTUs and
their sequences, a list of query sequences to assign to the reference OTUs, and
the sequence pairs that are within the distance threshold (e.g. 0.03) 
(Figure \ref{fig:algorithm}).
Initially, all query sequences are placed into separate OTUs. 
Then, the algorithm iteratively reassigns the query sequences to the reference 
OTUs to optimize the MCC. 
Alternatively, a sequence will remain unassigned if the MCC value is maximized
when the sequence is a singleton rather than assigned to a reference OTU.
All query and reference sequence pairs are considered when calculating the MCC.
This process is repeated until the MCC changes by no more than 0.0001 (default)
or until a maximum number of iterations is reached (default: 100).
In the closed reference mode, any query sequences that cannot be assigned to
references OTUs are discarded, and the results will only contain OTUs that exist
in the original reference.
In the open reference mode, unassigned query sequences are clustered _de novo_
using OptiClust to generate new OTUs.
The final MCC is reported with the best OTU assignments.
There are two strategies for generating OTUs with OptiFit: 
1) fit the query sequences to reference OTUs generated by _de novo_ clustering an
independent database, or
2) split the dataset into a reference and query fraction, cluster the reference
sequences _de novo_, then fit the query sequences to the reference OTUs.
We clustered sequences from four datasets isolated from soil
[@johnston_metagenomics_2016], marine [@henson_artificial_2016], mouse gut
[@schloss_stabilization_2012], and human gut [@baxter_microbiota-based_2016]
samples to test the performance of OptiFit with both of these strategies.

@@

While _de novo_ clustering produces high quality OTUs, researchers may prefer to
perform reference clustering to a public database because reference-based
methods produce consistent OTUs and are generally faster than _de novo_ methods.
In closed reference mode, sequences that cannot be assigned to reference OTUs
are thrown out, so that the final clustering contains only OTUs that exist in
the reference.
To test how OptiFit performs for this purpose, we fit each dataset to three
databases of reference OTUs: the Greengenes database, the SILVA non-redundant
database, and the Ribosomal Database Project (RDP) [@desantis_greengenes_2006;
@quast_silva_2013; @cole_ribosomal_2014].
Reference OTUs for each database were created by performing _de novo_ clustering
with OptiClust at a distance threshold of 3% (see Figure \ref{fig:workflow}).
The _de novo_ MCC scores were `r dn_dbs_mcc` for `r db_names_upper`,
respectively.
Fitting sequences to Greengenes and SILVA in closed reference mode performed
similarly, with median MCC scores of `r closed_fit_gg_mcc` and 
`r closed_fit_silva_mcc` respectively, while the median MCC 
was `r closed_fit_rdp_mcc` when fitting to RDP 
(see Figure \ref{fig:results_sum}).
For comparison, clustering datasets with OptiClust produced an average MCC score
of `r opticlust_mcc`.
This gap in OTU quality mostly disappeared when clustering in open reference
mode, which produced median MCCs of `r open_fit_gg_mcc` with Greengenes, 
`r open_fit_silva_mcc` with SILVA, and `r open_fit_rdp_mcc` with the RDP.
Thus, open reference OptiFit produced OTUs of very similar quality as _de novo_
clustering, and closed reference OptiFit followed closely behind as long as a
suitable reference database was chosen.

@@

When performing reference clustering against public databases, the database 
chosen greatly affects the quality of OTUs produced.
OTU quality may be poor when the reference database is too unrelated to the
samples of interest, such as when samples contain low abundant or novel
populations.
While _de novo_ clustering overcomes the quality limitations of reference
clustering to databases, OTU assignments are not consistent when new sequences
are added.
Researchers may wish to fit new sequences to existing OTUs when comparing OTUs
across studies or when making predictions with machine learning models.
To determine how well OptiFit performs for fitting new sequences to existing
OTUs, we employed a split dataset strategy, where each dataset was randomly
split into a reference fraction and a query fraction.
Reference sequences were clustered _de novo_ with OptiClust, then query
sequences were fit to the _de novo_ OTUs with OptiFit.

First, we tested whether OptiFit performed as well as _de novo_ clustering when
using the split dataset strategy with half of the sequences selected for
the reference by a simple random sample (a 50% split).
OTU quality was highly similar to that from OptiClust regardless of
mode (`r mcc_opticlust_vs_fit_split_simple`% difference in median MCC).
In closed reference mode, OptiFit was able to fit `r frac_fit_split`% of query
sequences to reference OTUs with the split strategy, a great improvement over
the average `r frac_fit_gg`% of sequences fit to the greengenes database.
In terms of run time, closed and open reference OptiFit performed faster than
OptiClust on whole datasets by `r sec_closed_fit_split_vs_clust`% and 
`r sec_open_fit_split_vs_clust` respectively.
The split dataset strategy also performed `r sec_closed_fit_split_vs_db`% faster
than the database strategy in closed reference mode and 
`r sec_open_fit_split_vs_db`% faster in open reference mode.
Thus, reference clustering with the split dataset strategy creates as high
quality OTUs as _de novo_ clustering yet at a faster run time, and fits far more
query sequences than the database strategy.

@@

While we initially tested this strategy using a 50% split of the data into
reference and query fractions, we next investigated whether there was
an optimal reference fraction size.
To test the best reference size, reference sets with 10% to 90% of the
sequences were created, with the remaining sequences used for the query.
OTU quality was remarkably consistent across reference fraction sizes.
For example, splitting the human dataset 100 times yielded a coefficient of
variation of `r cv_fit_split_mcc_human_simple` for the MCC score across all
fractions.
Run time generally decreased as the reference fraction increased; for the human
dataset, the median run time was `r sec_fit_split_human_simple_1` with 10% of
sequences in the reference and `r sec_fit_split_human_simple_9` with 90% of
sequences in the reference (Figure \ref{fig:results_split}).
In closed reference mode, the fraction of sequences that mapped increased as the
reference size increased; for the human dataset, the median fraction mapped was
`r frac_fit_split_human_simple_1` with 10% of sequences in the reference and 
`r frac_fit_split_human_simple_9` with 90% of sequences in the reference. 
These trends held for the other datasets as well (Figure 
\ref{fig:results_split}).
Thus, the reference fraction doid not affect OTU quality in terms of MCC score,
but did affect the run time and the fraction of sequences that mapped during the
closed reference clustering.

After testing the split strategy using a simple random sample to select the
reference sequences, we then investigated other methods of splitting the data.
We tested three methods for selecting the fraction of sequences to be used as
the reference at a size of 50%: a simple random sample, weighting sequences by
relative abundance, and weighting by similarity to other sequences in the
dataset.
OTU quality in terms of MCC was similar with the simple and abundance-weighted
sampling (median MCCs of `r mcc_fit_split_simple` and `r mcc_fit_split_abun`
respectively), but worse for similarity-weighted sampling (median MCC of
`r mcc_fit_split_dist`).
In closed-reference clustering mode, the fraction of sequences that mapped were
similar for simple and abundance-weighted sampling (median fraction mapped of 
`r frac_fit_split_simple` and `r frac_fit_split_abun` respectively), but worse 
for similarity-weighted sampling (median fraction mapped of `r frac_fit_split_dist`).
While simple and abundance-weighted sampling produced better quality OTUs than
similarity-weighted sampling, OptiFit performed faster on
similarity-weighted samples with a median runtime of `r sec_fit_split_dist`
seconds compared to `r sec_fit_split_simple` and `r sec_fit_split_abun` seconds
for simple and abundance-weighted sampling, respectively.
Thus, employing more complicated sampling strategies such as abundance-weighted
and similarity-weighted sampling did not confer any advantages over selecting
the reference via a simple random sample, and in fact decreased OTU quality in
the case of similarity-weighted sampling.

@@

When fitting to public databases, OTU quality dropped in closed reference mode
to different degrees depending on the database and dataset source, and no more
than half of query sequences were able to be fit to OTUs across any
dataset/database combination. 
This may reflect limitations of reference
databases, which are unlikely to contain sequences from rare and novel microbes.
This drop in quality was most notable with the RDP reference, which contained only 
`r ref_seqs[['rdp']]` sequences compared to `r ref_seqs[['silva']]` sequences in
SILVA and `r ref_seqs[['gg']]` in Greengenes after trimming to the V4 region.
Note that Greengenes has not been updated since 2013 at the time of this
writing, while SILVA is updated every one to two years with the most recent
release in 2019.
We recommend that users who require an independent reference database opt for
large databases with regular updates and good coverage of microbial diversity.
Since OptiClust still performs faster than open reference OptiFit and creates
higher quality OTUs than closed reference OptiFit with the database strategy, we
recommend using OptiClust rather than fitting to a database whenever consistent
OTUs are not required for the study at hand.

The OptiClust and OptiFit algorithms produced higher quality
OTUs than VSEARCH in open reference, closed reference, or _de novo_ modes.
However, VSEARCH was able to map more sequences to OTUs than OptiFit in closed
reference mode.
While both mothur and VSEARCH use a distance or similarity threshold for
determining how to assign sequences to OTUs, VSEARCH is more permissive than
mothur.
The OptiFit and OptiClust algorithms use all of the sequences to define an
OTU, requiring that all pairs of sequences (including reference and query
sequences) in an OTU are within the distance threshold without penalizing the
MCC.
In contrast, VSEARCH only requires each query sequence to be similar to the
single centroid sequence that seeded the OTU.
Additionally, OTU assignments clustered by VSEARCH are dependent on the order of
the input sequences, because each query is assigned to the OTU of the first
centroid sequence that is found within the distance threshold.
In this way, VSEARCH sacrifices OTU quality in order to allow more sequences to
fit to OTUs.

@@

We downloaded 16S rRNA gene amplicon sequences from four published datasets
isolated from soil [@johnston_metagenomics_2016], marine
[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
gut [@baxter_microbiota-based_2016] samples.
These datasets represent a selection of the broad types of natural communities
that microbial ecologists study.
We processed the raw sequences using mothur according to the Schloss Lab MiSeq
SOP as described in the mothur wiki and accompanying study by Kozich _et al._
[@schloss_miseq_nodate; @kozich_development_2013].
These steps included trimming and filtering for quality, aligning to the SILVA
reference alignment [@quast_silva_2013], discarding sequences that aligned
outside the V4 region, removing chimeric reads with UCHIME [@edgar_uchime_2011],
and calculating distances between all pairs of sequences within each dataset
prior to clustering.

>>>>

<!-- TODO
- Go back through https://peerj.com/articles/1487/ to get all the arguments for why the qiime based methods are problematic…
1. The threshold with closed reference is a radius and with de novo its a diameter
2. The gg reference was defined for full length sequences, within a subregion there are duplicate and nearly duplicate sequences
3. Because VSEARCH/USEARCH is a greedy heuristic, its looking for a reference that is close. Although the ordering of the database is not relevant (it was with USEARCH) it gets the wrong number of its (see Figure 5)
-->

<!-- TODO
- "require" and "enforce" for opti -- wrong word choice.
- be consistent with cluster, assign, fit, and map.
-->

@@

Assigning amplicon sequences to operational taxonomic units (OTUs) is often an
important step in characterizing the composition of microbial communities across
large datasets.
OptiClust, a _de novo_ OTU clustering method, has been
shown to produce higher quality OTU assignments than other methods and at
comparable or faster speeds.
A notable difference between _de novo_ clustering and database-dependent methods
is that OTU assignments from _de novo_ methods may change when new
sequences are added to a dataset.
However, in some cases one may wish to incorporate new samples into a previously
clustered dataset without performing clustering again on all sequences, such as
when comparing across datasets or deploying machine learning models where OTUs
are features.
**TODO**: sentence mentioning open/closed reference clustering and the inherent problems with the current methods.
To provide an efficient and robust method to fit amplicon sequence data to
existing OTUs, we developed the OptiFit algorithm.
We tested OptiFit using four microbiome datasets with two different strategies:
by fitting to an external reference database or by splitting the dataset into a 
reference and query set and fitting the query sequences to the reference set
after clustering it using OptiClust.
The result is a robust implementation of closed and open-reference clustering.
OptiFit produces OTUs of similar quality as OptiClust and at faster speeds when
using the split dataset strategy, although the OTU quality and processing speed
depends on the database chosen when using the external database strategy.
OptiFit provides a suitable option for users who require consistent OTU
assignments at the same quality afforded by _de novo_ clustering methods.

@@

Advancements in DNA sequencing technology have allowed researchers to affordably
generate millions of reads from microorganisms in diverse natural communities.
Efficient and robust software tools are needed to assign microbial sequences
into taxonomic groups for characterization and comparison of communities.
The OptiClust algorithm produces high quality groups by comparing sequences to
each other, but the assignments can change when new sequences are
added to a dataset, making it difficult to compare different studies.
Other approaches assign sequences to groups by comparing them to
sequences in a reference database to produce consistent assignments, but the
quality of the groups produced is reduced compared to OptiClust.
We developed OptiFit, a new reference-based algorithm that produces consistent
yet high quality assignments like OptiClust.
OptiFit allows researchers to compare microbial communities across different
studies or add new data to existing studies without sacrificing the quality of
the group assignments.

\newpage

@@

Amplicon sequencing is a mainstay of microbial ecology.
Researchers can affordably generate millions of sequences to characterize the
composition of hundreds of samples from microbial communities without the need
for culturing.
In many analysis pipelines, 16S rRNA gene sequences are assigned to
operational taxonomic units (OTUs) to facilitate comparison of taxonomic
composition between communities to avoid the need for taxonomic classification.
A distance threshold of 3% (or sequence similarity of 97%) is commonly used to
cluster sequences into OTUs based on pairwise comparisons of the sequences
within the dataset.
The method chosen for clustering affects the quality of OTU assignments and thus
may impact downstream analyses of community composition
[@westcott_opticlust_2017; @schloss_application_2016;
@westcott_novo_2015].

There are two main categories of OTU clustering algorithms: _de novo_ and
reference-based.
OptiClust is a _de novo_ clustering algorithm which uses the distance score
between all pairs of sequences in the dataset to cluster them into OTUs by
maximizing the Matthews Correlation Coefficient (MCC)
[@westcott_opticlust_2017].
This approach takes into account the distances between all pairs of sequences
when assigning query sequences to OTUs, in contrast to other _de novo_
methods such as the greedy clustering algorithms implemented in USEARCH and
VSEARCH, which only consider the distance between the query sequence and a
representative centroid sequence in the OTU [@edgar_search_2010;
@rognes_vsearch_2016].
In methods employing greedy clustering algorithms, some pairs of sequences in
the same OTU may have a greater distance than the specified threshold since only
the distance between each sequence and the centroid sequence is considered while
clustering.
In contrast, the OptiClust algorithm enforces that all pairs of sequences must
be within the distance threshold. <!-- i.e. radius vs. diameter -->
A limitation of _de novo_ clustering is that different OTU assignments will be
produced when new sequences are added to a dataset, making it difficult to use
_de novo_ clustering to compare OTUs between different studies.
Additionally, the greedy clustering algorithms are sensitive to the order of the
input sequences: different OTU assignments are produced when the same sequences
are randomly shuffled [@westcott_novo_2015; @he_stability_2015].
Furthermore, since _de novo_ clustering requires calculating and comparing
distances between all sequences in a dataset, the execution time can be slow for
very large datasets.
Reference clustering attempts to overcome the limitations of _de novo_
clustering methods by using a representative set of sequences from a database,
with each reference sequence seeding an OTU.
Commonly, the Greengenes set of representative full length sequences clustered
at 97% similarity is used as the reference with VSEARCH
[@desantis_greengenes_2006; @rognes_vsearch_2016; @noauthor_clustering_nodate].
Query sequences are then assigned to OTUs based on their similarity to the
reference sequences.
Any query sequences that are not within the distance threshold to any of the
reference sequences are either thrown out (closed reference clustering) or
clustered _de novo_ to create additional OTUs (open reference clustering).
While reference-based clustering is generally fast, it is limited by the
diversity of the reference database.
Rare or novel sequences in the sample will be lost in closed reference mode if
they are not represented by a similar sequence in the database.
Previous studies found that the OptiClust _de novo_ clustering algorithm created
the highest quality OTU assignments of all clustering methods [@westcott_opticlust_2017].

To overcome the limitations of current reference-based and _de novo_ clustering
algorithms while maintaining OTU quality, we developed OptiFit, a
reference-based clustering algorithm.
While other tools represent reference OTUs with a single sequence, OptiFit uses
multiple sequences in existing OTUs as the reference and fits new sequences
those reference OTUs.
In contrast to other tools, OptiFit considers all pairwise distance scores
between reference and query sequences when assigning sequences to OTUs in order
to produce OTUs of the highest possible quality.
Here, we tested the OptiFit algorithm with the reference as a public database
(e.g. Greengenes) or _de novo_ OTUs and compared the performance to existing
tools.
To evaluate the OptiFit algorithm and compare to existing methods, we used four
published datasets isolated from soil [@johnston_metagenomics_2016], marine
[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
gut [@baxter_microbiota-based_2016] samples. 
OptiFit is available within the mothur software program.

@@

OptiFit begins where OptiClust ends, starting with a list of reference OTUs and
their sequences, a list of query sequences to assign to the reference OTUs, and
the sequence pairs that are within the distance threshold (e.g. 0.03) 
(Figure \ref{fig:algorithm}).
Initially, all query sequences are placed into separate OTUs. 
Then, the algorithm iteratively reassigns the query sequences to the reference 
OTUs to optimize the MCC. 
Alternatively, a sequence will remain unassigned if the MCC value is maximized
when the sequence is a singleton rather than assigned to a reference OTU.
All query and reference sequence pairs are considered when calculating the MCC.
This process is repeated until the MCC changes by no more than 0.0001 (default)
or until a maximum number of iterations is reached (default: 100).
In the closed reference mode, any query sequences that cannot be assigned to
reference OTUs are discarded, and the results only contain OTUs that exist
in the original reference.
In the open reference mode, unassigned query sequences are clustered _de novo_
using OptiClust to generate new OTUs.
The final MCC is reported with the best OTU assignments.
There are two strategies for generating OTUs with OptiFit: 
1) fit the query sequences to reference OTUs generated by _de novo_ clustering an
independent database, or
2) split the dataset into a reference and query fraction, cluster the reference
sequences _de novo_, then fit the query sequences to the reference OTUs.

@@

While _de novo_ clustering produces the highest quality OTUs, researchers may
prefer to perform reference clustering to a public database because
reference-based methods produce consistent OTUs and are generally faster than
_de novo_ methods.
In closed reference mode, sequences that cannot be assigned to reference OTUs
are thrown out, so that the final clustering contains only OTUs that exist in
the reference.
To test how OptiFit performs for this purpose, we fit each dataset to three
databases of reference OTUs: the Greengenes database, the SILVA non-redundant
database, and the Ribosomal Database Project (RDP) [@desantis_greengenes_2006;
@quast_silva_2013; @cole_ribosomal_2014].
Reference OTUs for each database were created by performing _de novo_ clustering
with OptiClust at a distance threshold of 3% using the V4 region of each
sequence (see Figure \ref{fig:workflow}).
The _de novo_ MCC scores were `r dn_dbs_mcc` for `r db_names_upper`,
respectively.
Fitting sequences to Greengenes and SILVA in closed reference mode performed
similarly, with median MCC scores of `r closed_fit_gg_mcc` and 
`r closed_fit_silva_mcc` respectively, while the median MCC 
was `r closed_fit_rdp_mcc` when fitting to RDP 
(see Figure \ref{fig:results_sum}).
For comparison, clustering datasets with OptiClust produced an average MCC score
of `r opticlust_mcc`.
This gap in OTU quality mostly disappeared when clustering in open reference
mode, which produced median MCCs of `r open_fit_gg_mcc` with Greengenes, 
`r open_fit_silva_mcc` with SILVA, and `r open_fit_rdp_mcc` with the RDP.
Thus, open reference OptiFit produced OTUs of very similar quality as _de novo_
clustering, and closed reference OptiFit followed closely behind as long as a
suitable reference database was chosen.

@@

When performing reference clustering against public databases, the database 
chosen greatly affects the quality of OTUs produced.
OTU quality may be poor when the reference database consists of sequences that
are too unrelated to the samples of interest, such as when samples contain novel
populations.
While _de novo_ clustering overcomes the quality limitations of reference
clustering to databases, OTU assignments are not consistent when new sequences
are added.
Researchers may wish to fit new sequences to existing OTUs or to compare OTUs
across studies.
To determine how well OptiFit performs for fitting new sequences to existing
OTUs, we employed a split dataset strategy, where each dataset was randomly
split into a reference fraction and a query fraction.
Reference sequences were clustered _de novo_ with OptiClust, then query
sequences were fit to the _de novo_ OTUs with OptiFit.

First, we tested whether OptiFit performed as well as _de novo_ clustering when
using the split dataset strategy with half of the sequences selected for
the reference by a simple random sample (a 50% split).
OTU quality was highly similar to that from OptiClust regardless of
mode (`r mcc_opticlust_vs_fit_split_simple`% difference in median MCC).
In closed reference mode, OptiFit was able to fit `r frac_fit_split`% of query
sequences to reference OTUs with the split strategy, a great improvement over
the average `r frac_fit_gg`% of sequences fit to the Greengenes database.
In terms of run time, closed and open reference OptiFit performed faster than
OptiClust on whole datasets by `r sec_closed_fit_split_vs_clust`% and 
`r sec_open_fit_split_vs_clust` respectively.
The split dataset strategy also performed `r sec_closed_fit_split_vs_db`% faster
than the database strategy in closed reference mode and 
`r sec_open_fit_split_vs_db`% faster in open reference mode.
Thus, reference clustering with the split dataset strategy creates as high
quality OTUs as _de novo_ clustering yet at a faster run time, and fits far more
query sequences than the database strategy.

@@

While we initially tested this strategy using a 50% split of the data into
reference and query fractions, we next investigated whether there was
an optimal reference fraction size.
To identify the best reference size, reference sets with 10% to 90% of the
sequences were created, with the remaining sequences used for the query.
OTU quality was remarkably consistent across reference fraction sizes.
For example, splitting the human dataset 100 times yielded a coefficient of
variation of `r cv_fit_split_mcc_human_simple` for the MCC score across all
fractions.
Run time generally decreased as the reference fraction increased; for the human
dataset, the median run time was `r sec_fit_split_human_simple_1` with 10% of
sequences in the reference and `r sec_fit_split_human_simple_9` with 90% of
sequences in the reference (Figure \ref{fig:results_split}).
In closed reference mode, the fraction of sequences that mapped increased as the
reference size increased; for the human dataset, the median fraction mapped was
`r frac_fit_split_human_simple_1` with 10% of sequences in the reference and 
`r frac_fit_split_human_simple_9` with 90% of sequences in the reference. 
These trends held for the other datasets as well (Figure 
\ref{fig:results_split}).
Thus, the reference fraction did not affect OTU quality in terms of MCC score,
but did affect the run time and the fraction of sequences that mapped during the
closed reference clustering.

After testing the split strategy using a simple random sample to select the
reference sequences, we then investigated other methods of splitting the data.
We tested three methods for selecting the fraction of sequences to be used as
the reference at a size of 50%: a simple random sample, weighting sequences by
relative abundance, and weighting by similarity to other sequences in the
dataset.
OTU quality in terms of MCC was similar with the simple and abundance-weighted
sampling (median MCCs of `r mcc_fit_split_simple` and `r mcc_fit_split_abun`,
respectively), but worse for similarity-weighted sampling (median MCC of
`r mcc_fit_split_dist`).
In closed-reference clustering mode, the fraction of sequences that mapped were
similar for simple and abundance-weighted sampling (median fraction mapped of 
`r frac_fit_split_simple` and `r frac_fit_split_abun` respectively), but worse 
for similarity-weighted sampling (median fraction mapped of `r frac_fit_split_dist`).
While simple and abundance-weighted sampling produced better quality OTUs than
similarity-weighted sampling, OptiFit performed faster on
similarity-weighted samples with a median runtime of `r sec_fit_split_dist`
seconds compared to `r sec_fit_split_simple` and `r sec_fit_split_abun` seconds
for simple and abundance-weighted sampling, respectively.
Thus, employing more complicated sampling strategies such as abundance-weighted
and similarity-weighted sampling did not confer any advantages over selecting
the reference via a simple random sample, and in fact decreased OTU quality in
the case of similarity-weighted sampling.

@@

When fitting to public databases, OTU quality dropped in closed reference mode
to different degrees depending on the database and dataset source, and no more
than half of query sequences were able to be fit to OTUs across any
dataset/database combination. 
This may reflect limitations of reference databases, which are unlikely to
contain sequences from novel microbes.
This drop in quality was most notable with the RDP reference, which contained only 
`r ref_seqs[['rdp']]` sequences compared to `r ref_seqs[['silva']]` sequences in
SILVA and `r ref_seqs[['gg']]` in Greengenes.
Note that Greengenes has not been updated since 2013 at the time of this
writing, while SILVA and the RDP are updated regularly.
We recommend that users who require an independent reference database opt for
large databases with regular updates and good coverage of microbial diversity
for their environment.
Since OptiClust still performs faster than open reference OptiFit and creates
higher quality OTUs than closed reference OptiFit with the database strategy, we
recommend using OptiClust rather than fitting to a database whenever consistent
OTUs are not required.

The OptiClust and OptiFit algorithms produced higher quality
OTUs than VSEARCH in open reference, closed reference, or _de novo_ modes.
However, VSEARCH was able to map more sequences to OTUs than OptiFit in closed
reference mode.
While both OptiFit and VSEARCH use a distance or similarity threshold for
determining how to assign sequences to OTUs, VSEARCH is more permissive than
OptiFit.
The OptiFit and OptiClust algorithms use all of the sequences to define an
OTU, requiring that all pairs of sequences (including reference and query
sequences) in an OTU are within the distance threshold without penalizing the
MCC.
In contrast, VSEARCH only requires each query sequence to be similar to the
single centroid sequence that seeded the OTU.
Additionally, OTU assignments clustered by VSEARCH are dependent on the order of
the input sequences, because each query is assigned to the OTU of the first
centroid sequence that is found within the distance threshold. TODO: is this true of VSEARCH or only USEARCH?
Because of this, VSEARCH sacrifices OTU quality by allowing more sequences to
fit to OTUs.

@@

We downloaded 16S rRNA gene amplicon sequences from four published datasets
isolated from soil [@johnston_metagenomics_2016], marine
[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
gut [@baxter_microbiota-based_2016] samples.
These datasets contain sequences from the V4 region of the 16S rRNA gene and
represent a selection of the broad types of natural communities that microbial
ecologists study.
We processed the raw sequences using mothur according to the Schloss Lab MiSeq
SOP and accompanying study by Kozich _et al._
[@schloss_miseq_nodate; @kozich_development_2013].
These steps included trimming and filtering for quality, aligning to the SILVA
reference alignment [@quast_silva_2013], discarding sequences that aligned
outside the V4 region, removing chimeric reads with UCHIME [@edgar_uchime_2011],
and calculating distances between all pairs of sequences within each dataset
prior to clustering.
